---
title: "Графики к lec 08"
author: "Boris Demeshev"
output:
  pdf_document:
    keep_tex: yes
lang: russian
header-includes:
   - \usepackage{dcolumn}
   - \usepackage{embedfile}
---



```{r, include=FALSE}
library("knitr")
library("tikzsetup")
tikzsetup()
opts_chunk$set(echo=FALSE)
```


```{r}
library("ggplot2")
library("forecast")
library("sophisthse")
library("pander")
library("dplyr")
n <- 200
```

Страница 2/62. Таблица
```{r}
y <- sophisthse(series.name = c("POPNUM_Y","GDPVA_EEA_Y"), output="data.frame")
y_sel <- y %>% filter(T>2001,T<2014) %>% select(T, POPNUM_Y, GDPVA_EEA_Y)
colnames(y_sel) <- c("Год", "Население (тыс. чел.)", "ВВП (млрд. руб.)")
pander(y_sel)
```





\newpage
Страница 9/62

```{r}
y <- arima.sim(n=n, list(order=c(0,0,0)), sd = 2 )

# tsdisplay(y)
plot(y,main = "Белый шум, $y_t=\\varepsilon_t \\sim N(0,4)$")
```


\newpage
Страница 12/62

```{r}
y <- 5+6*(1:n) + arima.sim(n=n, list(order=c(0,0,0)), sd=200 )

# tsdisplay(y)
plot(y,main = "Процесс с трендом, $y_t=5+6t+\\varepsilon_t$, $\\varepsilon_t\\sim N(0,200^2)$")
```



\newpage
Страница 13/62

```{r}
y <- arima.sim(n=n, list(order=c(0,0,0)) )

# tsdisplay(y)
plot(y,main = "Случайное блуждание, $y_t=y_{t-1}+ 2 + \\varepsilon_t$")
```



\newpage
Страница 56/62

```{r}
y <- arima.sim(n=n, list(order=c(0,0,0)) )

# tsdisplay(y)
plot(y,main = "Белый шум, $y_t=\\varepsilon_t$")
```

\newpage
Страница 56/62

```{r}
Acf(y,main = "ACF, $y_t=\\varepsilon_t$")
```

\newpage
Страница 56/62

```{r}
Pacf(y,main = "PACF, $y_t=\\varepsilon_t$")
```

\newpage
Страница 57/62


```{r}
y <- arima.sim(n=n, list(order=c(0,1,0)) )
plot(y,main = "Случайное блуждание, $y_t=y_{t-1}+\\varepsilon_t$")
```

\newpage
Страница 57/62

```{r}
Acf(y,main = "ACF, $y_t=y_{t-1}+\\varepsilon_t$")
```

\newpage
Страница 57/62

```{r}
Pacf(y,main = "PACF, $y_t=y_{t-1}+\\varepsilon_t$")
```

\newpage
Страница 58/62


```{r}
y <- 0.02*(1:n)+arima.sim(n=n, list(order=c(0,0,0)) )
plot(y,main = "Процесс с трендом, $y_t=0.02\\cdot t +\\varepsilon_t$")

```

\newpage
Страница 58/62

```{r}
Acf(y,main = "ACF, $y_t=0.02\\cdot t +\\varepsilon_t$")
```

\newpage
Страница 58/62

```{r}
Pacf(y,main = "PACF, $y_t=0.02\\cdot t +\\varepsilon_t$")
```


\newpage
Страница 59/62


```{r}
y <- arima.sim(n=n, list(ar=0.7) )
plot(y,main = "AR(1), $y_t=0.7y_{t-1} +\\varepsilon_t$")
```

\newpage
Страница 59/62

```{r}
Acf(y, main = "ACF, AR(1), $y_t=0.7y_{t-1}+\\varepsilon_t$")
```

\newpage
Страница 59/62

```{r}
Pacf(y, main = "PACF, AR(1), $y_t=0.7y_{t-1} +\\varepsilon_t$")
```

\newpage
Страница 59/62

```{r}
y <- arima.sim(n=n, list(ar=c(0.9,-0.5)) )
plot(y,main = "AR(2), $y_t=0.9y_{t-1}-0.5y_{t-2} +\\varepsilon_t$")
```

\newpage
Страница 59/62

```{r}
Acf(y, main = "ACF, AR(2), $y_t=0.9y_{t-1}-0.5y_{t-2} +\\varepsilon_t$")
```

\newpage
Страница 59/62

```{r}
Pacf(y, main = "PACF, AR(2), $y_t=0.9y_{t-1}-0.5y_{t-2} +\\varepsilon_t$")
```


\newpage
Страница 60/62


```{r}
y <- arima.sim(n=n, list(ma=0.7) )
plot(y,main = "MA(1), $y_t=0.7\\varepsilon_{t-1} +\\varepsilon_t$")
```

\newpage
Страница 60/62

```{r}
Acf(y, main = "ACF, MA(1), $y_t=0.7\\varepsilon_{t-1} +\\varepsilon_t$")
```

\newpage
Страница 60/62

```{r}
Pacf(y, main = "PACF, MA(1), $y_t=0.7\\varepsilon_{t-1} +\\varepsilon_t$")
```

\newpage
Страница 60/62

```{r}
y <- arima.sim(n=n, list(ma=c(0.9,0.5)) )
plot(y,main = "MA(2), $y_t=0.9\\varepsilon_{t-1} +0.5\\varepsilon_{t-1}+\\varepsilon_t$")
```

\newpage
Страница 60/62

```{r}
Acf(y, main = "ACF, MA(2), $y_t=0.9\\varepsilon_{t-1} +0.5\\varepsilon_{t-1}+\\varepsilon_t$")
```

\newpage
Страница 60/62

```{r}
Pacf(y, main = "PACF, MA(2), $y_t=0.9\\varepsilon_{t-1} +0.5\\varepsilon_{t-1}+\\varepsilon_t$")
```

\newpage
Страница 61/62


```{r}
y <- arima.sim(n=n, list(ar=0.7, ma=0.5) )
plot(y,main = "ARMA(1,1), $y_t=0.7y_{t-1}+0.5\\varepsilon_{t-1} +\\varepsilon_t$")
```

\newpage
Страница 61/62

```{r}
Acf(y, main = "ACF, ARMA(1,1), $y_t=0.7y_{t-1}+0.5\\varepsilon_{t-1} +\\varepsilon_t$")
```

\newpage
Страница 61/62

```{r}
Pacf(y, main = "PACF, ARMA(1,1), $y_t=0.7y_{t-1}+0.5\\varepsilon_{t-1} +\\varepsilon_t$")
```
